---
output:
  word_document: default
  html_document: default
---
title: "Prediction Assignment for Practical Machine Learning Course"

author: "ITguy666"

date: "2026-06-20"

## Data Loading and Processing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(caret)
library(randomForest)
library(gbm)
library(data.table)
library(dplyr)

# Download files
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",destfile="training.csv",method = "curl")
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",destfile="testing.csv",method = "curl")

# Read directly from the web to bypass local file errors
training1 <- read.csv("training.csv",na.strings=c('#DIV/0!', '', 'NA'))
testing1  <- read.csv("testing.csv",na.strings=c('#DIV/0!', '', 'NA'))

# Check dimensions to confirm successful loading
dim(training1)
```

## Data Cleaning & Variable Selection

To prevent overfitting and reduce training time, we remove irrelevant features. Columns with more than 95% NA values will be removed, as well as near-zero variance predictors.

```{r}
#Remove Near-Zero Variance predictors
nzv_metrics <- nearZeroVar(training1, saveMetrics = TRUE)
training2 <- training1[, nzv_metrics$nzv == FALSE]

# Remove columns with vast majority missing values
na_threshold <- sapply(training2, function(x) mean(is.na(x))) > 0.95
training2 <- training2[, na_threshold == FALSE]

# Exclude metadata and descriptive identifiers as the first 7 columns store time stamps, 
# user names, and window sequences. These are not relevant for physical activity telemetry.
training2 <- training2[, -(1:7)]
training2$classe <- as.factor(training2$classe)
dim(training2)
```

With the cleaned up training2 data featuring 53 structured predictors, we split the dataset further into an internal sub-training set (70%) and an internal validation set (30%).

```{r}
in_train <- createDataPartition(y = training2$classe, p = 0.70, list = FALSE)
sub_train <- training2[in_train, ]
sub_validate <- training2[-in_train, ]
```

## Predictive Modelling

In our trainings, a 5-fold cross-validation strategy within the Caret train control wrapper was deployed to preserve computational resources while maintaining robust tuning metrics.

Two models are built for comparison:

Model 1: Random Forest (Naturally constructs multi-class decision trees and works incredibly well with high-dimensional sensor data).

Model 2: Gradient Boosting Machine (GBM) (Utilized as a baseline comparison model).

```{r}
fit_control <- trainControl(method = "cv", number = 5, verboseIter = FALSE)

# Model 1: Random Forest
model_rf <- train(classe ~ ., data = sub_train, method = "rf", trControl = fit_control)
print(model_rf$finalModel)

# Model 2: GBM
model_gbm <- train(classe ~ ., data = sub_train, method = "gbm", trControl = fit_control, verbose = FALSE)
print(model_gbm$finalModel)
```

## Model Evaluation & Out-of-Sample Error

We pass the internal validation matrix (sub_validate) through both models to generate accuracy and sample confusion matrix calculations.

```{r}
# Evaluate Random Forest Model 1
pred_rf <- predict(model_rf, sub_validate)
matrix_rf <- confusionMatrix(pred_rf, sub_validate$classe)
print(matrix_rf$overall['Accuracy'])

# Evaluate GBM Model 2
pred_gbm <- predict(model_gbm, sub_validate)
matrix_gbm <- confusionMatrix(pred_gbm, sub_validate$classe)
print(matrix_gbm$overall['Accuracy'])
```

## Final Model Selection & Expected Error

Comparing both setups, the Random Forest algorithm yielded a superior accuracy rate.Given the exceptionally low error rate, the Random Forest model is selected to compute the final quiz submission.

## Final Course Test Quiz Predictions

We load the raw course testing matrix, extract matching sensor attributes, and output the vector classification answers for the automated evaluation portal.

```{r}
clean_features <- colnames(training2[, -ncol(training2)])
final_testing   <- testing1[, clean_features]

quiz_predictions <- predict(model_rf, final_testing)
quiz_predictions
```
